Carbonate clumped isotope constraints on Silurian ocean temperature and seawater $\delta^{18}$O

### 2019-03-27 Silurian_ocean_temperature/draft.tex
Move files into Silurian_ocean_temperature directory

\documentclass[5p, authoryear]{elsarticle}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{natbib}
\usepackage{lineno}
\usepackage{dblfloatfix}
\usepackage[skip=2pt,font=footnotesize]{caption}
\usepackage{breakcites}

\renewcommand{\topfraction}{0.99}
\renewcommand\bottomfraction{.99}
\renewcommand{\textfraction}{0.01}
\renewcommand{\floatpagefraction}{0.99}
\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

@@

\title{Carbonate clumped isotope constraints on Silurian ocean temperature \\and seawater $\delta^{18}$O}

\author[caltech]{Renata C. Cummins\corref{cor1}}
\ead{renatacummins@gmail.com}

@@

Much of what we know about the history of Earth's climate derives from the chemistry of carbonate minerals in the sedimentary record. The oxygen isotopic composition ($\delta^{18}$O) of calcitic marine fossils and cements has been widely used as a proxy for past seawater temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the $\delta^{18}$O of seawater in the past. Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a calcite phase precipitated, allowing direct calculation of the $\delta^{18}$O of the water as well. The clumped isotope composition of calcites is also highly sensitive to and can help diagnose different modes of diagenetic alteration, enabling evaluation of preservation states and identification of the most pristine materials from within a sample set---critical information for evaluating the quality of paleoproxy data generated from carbonates.

We evaluated preservation and measured the clumped isotope composition of a suite of calcitic fossils (primarily brachiopods and corals), sedimentary grains, and cements from Silurian (ca. 433 Ma) stratigraphic sections on the island of Gotland, Sweden. Substantial variability in clumped isotope temperatures suggests differential preservation with alteration largely tied to rock-buffered diagenesis, complicating the generation of a stratigraphically resolved climate history through these sections. Despite the generally high preservation quality of samples from these sections, micro-scale observations of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to only a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm (31$\pm$2$^{\circ}$C) and similar in oxygen isotopic composition to that estimated for a ``modern'' ice-free world ($\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$). This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's ocean waters has remained broadly constant through time.

@@

Despite more than half a century of investigations on the topic, the long-term temperature history of Earth's oceans remains poorly constrained. The most widely used proxy for past ocean temperature is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate $\delta^{18}$O thermometer \citep{Urey1947}. This thermometer is based on the temperature dependence of the fractionation between the $\delta^{18}$O of carbonate, here reported relative to the VPDB standard, and the $\delta^{18}$O of seawater, here reported relative to the VSMOW standard. The carbonate $\delta^{18}$O proxy has two significant complications. First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary environmental signature. This is a fundamental challenge associated with carbonate minerals, which commonly undergo recrystallization during diagenesis and burial, altering primary geochemical signals and impacting the quality of proxy data. Second, the measured $\delta^{18}$O of carbonates depends not only on the temperature of the water from which the carbonates precipitated, but also on the $\delta^{18}$O of the fluid (seawater and/or pore fluids). The $\delta^{18}$O of seawater can be influenced by several processes, including local evaporation and mixing with meteoric water, as well as global changes in the volume and isotopic composition of continental glaciers, all of which are difficult to constrain independently in the geologic record.

Based on observations of carbonates, it was hypothesized that the $\delta^{18}$O of the ocean increased gradually but substantially through Earth's history---by $\sim$8$\permil$ over Phanerozoic time \citep{Jaffres2007, Veizer1999}. On 100-million-year timescales, gradual changes in $\delta^{18}$O are, in principle, plausible given current knowledge of the fluxes impacting the seawater $\delta^{18}$O budget. The long-term average $\delta^{18}$O of all the water on Earth's surface is largely controlled by the balance between two classes of processes within the silicate Earth: exchange during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange during low-temperature weathering of oceanic and continental crust \citep{Muehlenbachs1998}. High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the $\delta^{18}$O of ocean water by equilibrating the water with the average mantle $\delta^{18}$O$_{\text{VSMOW}}$ of 5.8$\permil$ \citep{Muehlenbachs1998}. Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the $\delta^{18}$O of seawater by preferentially moving $^{18}$O from water into rocks. Furthermore, we might reasonably anticipate that the relative magnitudes of high- and low-temperature weathering have changed over time with continental assembly and rifting \citep{Collins2003}, changes in seafloor spreading and production of oceanic crust \citep{Muller2008, Fornari1995}, climate \citep{Frakes2005}, and sedimentation \citep{Molnar2004}, unless a set of feedbacks, albeit enigmatic, operate to buffer the $\delta^{18}$O of seawater \citep{Gregory1981, Gregory1991, Muehlenbachs1976, Muehlenbachs1998, Jaffres2007}.

The most widely cited evidence for gradual change in seawater $\delta^{18}$O, however, remains a longstanding controversy in geochemistry. This arises from the complications associated with the carbonate $\delta^{18}$O proxy, which allow conflicting interpretations of the observed $\sim$8$\permil$ increase in the average $\delta^{18}$O of marine carbonates from the Cambrian to today \citep{Jaffres2007, Veizer1999}. Several studies have proposed that early Paleozoic carbonates may have low $\delta^{18}$O values because they are more likely to have undergone diagenetic alteration \citep{Clayton1959, Keith1964, Land1995}, which tends to decrease the $\delta^{18}$O of carbonate \citep{Marshall1992}. In response to this concern, \cite{Veizer1999} measured the $\delta^{18}$O of calcite fossils that were screened for diagenetic alteration using trace metal compositions and micro-scale studies of calcite fabrics. These screened fossils also record increasing $\delta^{18}$O through the Phanerozoic, arguing against diagenetic alteration as a cause of the low $\delta^{18}$O values in the early Paleozoic. An alternative hypothesis holds that oceans could have been much warmer in early Paleozoic time \citep{Karhu1986, Knauth1976}. Assuming that the $\delta^{18}$O of the oceans was the same as today, Ordovician tropical oceans must have been $\sim$45$^{\circ}$C to produce carbonates with the observed average $\delta^{18}$O$_{\text{VPDB}}$ of -5.9$\permil$ \citep{Veizer1999}. Such warm temperatures present an apparent paradox with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading several authors to suggest that the increase in the $\delta^{18}$O of marine carbonates must therefore reflect a concomitant increase in the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater from -8$\permil$ in the Cambrian to its present value of $\sim$0$\permil$ \citep{Jaffres2007, Veizer1999}.

Carbonate clumped isotope thermometry supplies a unique insight into this problem because it provides data that can help resolve the two main complications of the carbonate $\delta^{18}$O proxy: the clumped isotope composition of a carbonate is independent of the $\delta^{18}$O of water and highly sensitive to diagenetic alteration. The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates \citep{Eiler2011}, and isotopic clumping in a wide diversity of shallow-water calcifiers has been shown to record ambient water temperatures \citep{Ghosh2006, Henkes2013, Saenger2012, Tripati2010}. Clumped isotope temperature estimates can in turn be used, with the $\delta^{18}$O of shell calcite and the experimentally determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}, to calculate the $\delta^{18}$O of the water from which the calcite was precipitated.

Shells that have been deeply buried in lithified sediments have often experienced high-temperature alteration, which can change the $\delta^{18}$O of the carbonate and is frequently difficult to detect. Clumped isotopes are sensitive to this type of alteration because the altered material will take on a new clumped isotope temperature that can be much higher than the temperature of the pristine fossil calcite. If the altered material is incorporated within pristine fossil material selected for sample analysis (e.g. as cements lining fossil void space), the clumped isotope temperature of the sample will be significantly higher than the temperature of the ocean water where the organism lived. We demonstrate here that carbonate clumped isotopes can provide a more sensitive indicator of diagenetic alteration than the $\delta^{18}$O or the trace metal content of a carbonate.

In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotope arrays are also diagnostic of diagenetic regimes, allowing identification of the most pristine end-member of a sample set. Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and fluid-buffered diagenesis, respectively) and can also identify fluid-buffered diagenesis in the presence of meteoric water. As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. During rock-buffered diagenesis, this causes the $\delta^{18}$O of the water to increase, while the $\delta^{18}$O of the altered carbonate is not affected. In a water-buffered diagenetic regime, increasing temperature instead causes the $\delta^{18}$O of the altered carbonate to decrease. In contrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the $\delta^{18}$O of both the carbonate and the water are diagnostically low. The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}. We demonstrate here that this distinction has important implications for the estimation of the sea surface temperature and $\delta^{18}$O of seawater in the past.

\begin{figure*}[t]
\centering
\includegraphics[width=140mm]{Fig_01-01.png}
\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}. Locations of stratigraphic sections are marked and labeled with location number, which corresponds to the sample locations in Supplementary Table 1.}
\label{map}
\end{figure*}

@@

Samples were collected from several sections exposed on the island of Gotland, Sweden that span the latest Llandovery through Wenlock epochs of the Silurian Period (Figure \ref{map}). Ages of the formations and correlations between outcrops are constrained by conodont and graptolite biostratigraphy \citep{Jeppsson2006}. The sections record shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was constrained to be 19$^{\circ}$ south of the equator during the Silurian Period \citep{Torsvik1992}.

The oldest exposed strata on Gotland comprise the Lower Visby Formation, which consists of marls interbedded with fine-grained limestones deposited below storm wave base (Figure \ref{column}). Fossils are abundant throughout this unit \citep{Calner2004a, Samtleben1996}. The contact between the Lower and Upper Visby formations is marked by a condensed interval with abundant fossils of the large solitary rugose coral genus \textit{Phaulactis} \citep{Jeppsson1997, Jeppsson2006, Munnecke2003, Samtleben1996}. This contact occurs just above the Llandovery-Wenlock boundary as defined by conodont biostratigraphy \citep{Aldridge1993, Jeppsson1983, Mabillard1985}. We defined the \textit{Phaulactis} bed as a stratigraphic datum to align the sections, except those in the Slite Formation, which are measured relative to an arbitrary bed in the Slite Formation. The contact between the Lower and Upper Visby formations is also characterized by a facies change that represents a maximum flooding interval \citep{Calner2004b}. Carbonate content increases in the Upper Visby Beds, with a decreasing frequency of marl beds and the appearance of bioclastic limestones and reef mounds. Ripple bedforms and cross-stratification indicate deposition above storm wave base. The Upper Visby Formation is particularly fossiliferous \citep{Calner2004a, Samtleben1996}.

\begin{figure*}[t]
\centering
\includegraphics[width=140mm]{Fig_02-01.png}
\caption{Lithology and isotope stratigraphy of the sampled sequence. On the left, the stratigraphic column shows the generalized lithostratigraphic units sampled in this study. On the right are the carbon and oxygen isotopic compositions of brachiopods, collected from the same sections by \cite{Bickert1997}. Isotope data is from \cite{Bickert1997}.}
\label{column}
\end{figure*}

@@

The Silurian section on Gotland is an attractive location to sample for a clumped isotope study for several reasons.

First, this section has long been recognized for its generally excellent fossil preservation. The strata on Gotland are very nearly flat lying, with a dip of less than 1$^{\circ}$, display only minimal evidence of tectonic disturbance and have not been deeply buried \citep{Calner2004a, Jeppsson1983}. Conodont fossils from this section yield a conodont alteration index (CAI) of one, which suggests a maximum temperature of $\sim$100$^{\circ}$C \citep{Epstein1976, Jeppsson1983, Wenzel2000}.

The Silurian section on Gotland is also attractive because of the abundance of fossils, especially brachiopods, throughout the section. As with previous paleoclimate studies we sampled fossil phases for clumped isotope measurements in order to target calcite that was first precipitated directly from seawater (as opposed to carbonate matrix, which may contain authigenic cements). Many studies have targeted the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998, Samtleben2001}. The fibrous calcite fabric characteristic of the secondary layer has been well characterized in modern and fossil brachiopods \citep{PerezHuerta2007, Samtleben2001, Schmahl2004}, providing a standard of comparison for the samples in our dataset. The trace metal composition of brachiopods has also been extensively studied as an indicator of diagenetic alteration \citep{Azmy1998, Brand2003, Brand2012, Mii1994, Grossman1996, Shields2003}. The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often for paleoenvironmental studies than brachiopods, but have been examined by other clumped isotope studies of paleotemperature \citep{Came2007, Dennis2013, Finnegan2011}. Measurements of several fossil taxa allow comparison of the relative preservation of different calcitic materials and can help to detect taxon-specific ``vital effects'' if they are present.

A third attractive feature of the Silurian section on Gotland is the previously observed positive excursion in $\delta^{13}$C that initiates at the boundary between the Lower and Upper Visby Beds (Figure \ref{column}). This excursion, termed the Ireviken event, is now recognized globally \citep{Munnecke2003}. In Gotland, a 3$\permil$ excursion in $\delta^{13}$C is coupled to a poorly resolved 0.6$\permil$ positive excursion in $\delta^{18}$O \citep{Munnecke2003}. It is somewhat surprising that the excursions coincide with a maximum flooding interval, but it is difficult to distinguish the relative contributions of sediment supply and eustatic sea level rise to this facies change \citep{Calner2004a}. The excursions are also broadly coincident with minor extinctions in several benthic and pelagic marine taxa \citep{Munnecke2003}.

The coincident isotopic excursions and stratigraphic change suggest that the Ireviken Event captures a climate change phenomenon, the nature of which is still debated. The similarity between the isotopic signatures of the Ireviken event and known glaciations, such as the Late Ordovician glaciation, have led several studies to hypothesize a glacial cause for the Ireviken event \citep{Azmy1998, Kaljo2003, Brand2006, Calner2008}. This hypothesis was thought to be supported by the presence of glacial tillites in Brazil that were at first interpreted to be early Wenlock in age \citep{Grahn1992}, but these diamictites were later re-interpreted as re-workings of earlier glacial deposits. Early Silurian glaciation in South America is now thought to have ended prior to the late Llandovery, before the Ireviken Event \citep{DiazMartinez2007a, DiazMartinez2007b}. Many of these observations could also be explained by a shift from a humid period to an arid period in the tropics, which could accompany a glaciation or occur independently \citep{Bickert1997, Jeppsson1990, Munnecke2003, Samtleben1996}. According to this model, during the humid period recorded by the Lower Visby Beds, increased precipitation would have lowered the $\delta^{18}$O of surface waters and also stimulated the upwelling of deep water with low $\delta^{13}$C onto the shelves. Increased precipitation would also have caused intense weathering, which could explain the dominance of marls over carbonate reefs during this time. During the succeeding arid period beginning in the Upper Visby Beds, evaporation would have increased the $\delta^{18}$O of surface waters, and downwelling of saline water would have brought surface water with high $\delta^{13}$C onto the shelves. Low terrigenous influxes would have allowed the growth of carbonate reefs as observed in the H\"{o}gklint Formation. These hypotheses make different predictions for temperature and the oxygen isotopic composition of seawater. Here we test our ability to resolve possible changes in ocean temperature and/or seawater $\delta^{18}$O through the Ireviken event using stratigraphic samples from Gotland. Resolving a change this small with the uncertainties associated with carbonate clumped isotope thermometer is challenging but not impossible, because a shift of 0.6$\permil$ in $\delta^{18}$O corresponds to a change in temperature of approximately 2$^{\circ}$C. Though the uncertainties are commonly higher, single measurements of carbonate clumped isotope composition can attain a resolution of $\pm$1$^{\circ}$C, while running multiple replicates of a sample can improve the resolution to $\pm$0.4$^{\circ}$C \citep{Thiagarajan2011}.

>>>>

\documentclass[preprint, authoryear]{elsarticle}
\usepackage{amsmath} % required for basic math equations
\usepackage{wasysym} % required for permil symbol
\usepackage{natbib}

@@

\title{Carbonate clumped isotope constraints on Silurian ocean temperature and seawater $\delta^{18}$O}

\author[caltech]{Renata C. Cummins\corref{cor1}}
\ead{rcummins@post.harvard.edu}

@@

Much of what we know about the history of Earth's climate derives from the chemistry of carbonate minerals in the sedimentary record. The oxygen isotopic compositions ($\delta^{18}$O) of calcitic marine fossils and cements have been widely used as a proxy for past seawater temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the $\delta^{18}$O of seawater in the past. Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a calcite phase precipitated, and allows direct calculation of the $\delta^{18}$O of the water. The clumped isotope composition of calcites is also highly sensitive to recrystallization and can help diagnose different modes of diagenetic alteration, enabling evaluation of preservation states and identification of the most pristine materials from within a sample set---critical information for assessing the quality of paleoproxy data generated from carbonates.

We measured the clumped isotope composition of a large suite of calcitic fossils (primarily brachiopods and corals), sedimentary grains, and cements from Silurian (ca. 433 Ma) stratigraphic sections on the island of Gotland, Sweden. Substantial variability in clumped isotope temperatures suggests differential preservation with alteration largely tied to rock-buffered diagenesis, complicating the generation of a stratigraphically resolved climate history through these sections. Despite the generally high preservation quality of samples from these sections, micro-scale observations of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis suggest that only a subset of relatively pristine samples retain primary clumped isotope signatures. These samples indicate that Silurian tropical oceans were likely warm (31$\pm$2$^{\circ}$C) and similar in oxygen isotopic composition to that estimated for a ``modern'' ice-free world ($\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$). This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's ocean waters has remained broadly constant through time.

@@

Despite more than half a century of study, the long-term temperature history of Earth's oceans remains poorly constrained. The most widely used proxy for past ocean temperature is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate $\delta^{18}$O thermometer \citep{Urey1947}. This thermometer is based on the temperature dependence of the fractionation between the $\delta^{18}$O of carbonate, here reported relative to the VPDB standard, and the $\delta^{18}$O of seawater, here reported relative to the VSMOW standard. The carbonate $\delta^{18}$O proxy has two significant complications. First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from a primary environmental signature. This is an acute challenge associated with carbonate minerals, which commonly undergo recrystallization during diagenesis and burial, altering primary geochemical signals and impacting the quality of proxy data. Second, the measured $\delta^{18}$O of carbonates depends not only on the temperature of the water from which the carbonates precipitated, but also on the $\delta^{18}$O of the fluid (seawater and/or pore fluids). The $\delta^{18}$O of seawater can be influenced by several processes, including local evaporation and mixing with meteoric water, as well as global changes in the volume and isotopic composition of continental glaciers, all of which are difficult to constrain independently in the geologic record.

Based largely on observations of carbonates, it was hypothesized that the $\delta^{18}$O of the ocean increased gradually but substantially through Earth's history---by $\sim8\permil$ over Phanerozoic time \citep{Jaffres2007, Veizer1999}. On 100-million-year timescales, gradual changes in $\delta^{18}$O are, in principle, possible given current knowledge of the fluxes impacting the seawater $\delta^{18}$O budget. The long-term average $\delta^{18}$O of all the water on Earth's surface is largely controlled by the balance between two classes of processes within the silicate Earth: exchange during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange during low-temperature weathering of oceanic and continental crust \citep{Muehlenbachs1998}. High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the $\delta^{18}$O of ocean water by equilibrating the water with the average mantle $\delta^{18}$O$_{\text{VSMOW}}$ of 5.8$\permil$ \citep{Muehlenbachs1998}. Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the $\delta^{18}$O of seawater by preferentially moving $^{18}$O from water into rocks. Furthermore, we might reasonably anticipate that the relative magnitudes of high- and low-temperature weathering have changed over time with continental assembly and rifting \citep{Collins2003}, changes in seafloor spreading and production of oceanic crust \citep{Muller2008, Fornari1995}, climate \citep{Frakes2005}, and sedimentation \citep{Molnar2004}, unless a set of feedbacks, albeit enigmatic, operate to buffer the $\delta^{18}$O of seawater \citep{Gregory1981, Gregory1991, Muehlenbachs1976, Muehlenbachs1998, Jaffres2007}.

The most widely cited evidence for gradual change in seawater $\delta^{18}$O is an observed ca. 8$\permil$ increase in the average $\delta^{18}$O of marine carbonates from the Cambrian to today \citep{Jaffres2007, Veizer1999}. However, the complications associated with the carbonate $\delta^{18}$O proxy allow conflicting interpretations of this trend in carbonate $\delta^{18}$O, giving rise to a longstanding controversy. Several studies have proposed that early Paleozoic carbonates may have low $\delta^{18}$O values because they are more likely to have undergone diagenetic alteration \citep{Clayton1959, Keith1964, Land1995}, which tends to decrease the $\delta^{18}$O of carbonate \citep{Marshall1992}. In response to this concern, \cite{Veizer1999} measured the $\delta^{18}$O of calcite fossils that were screened for diagenetic alteration using trace metal compositions and micro-scale studies of calcite fabrics. These screened fossils also record increasing $\delta^{18}$O through the Phanerozoic, arguing against diagenetic alteration as a cause of the low $\delta^{18}$O values in the early Paleozoic. An alternative hypothesis holds that oceans were much warmer in early Paleozoic time \citep{Karhu1986, Knauth1976}. Assuming that the $\delta^{18}$O of the oceans was the same as today, Ordovician tropical oceans must have been $\sim$45$^{\circ}$C to produce carbonates with the observed average $\delta^{18}$O$_{\text{VPDB}}$ of -5.9$\permil$. Such warm temperature estimates present an apparent paradox with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading several authors to suggest that the increase in the $\delta^{18}$O of marine carbonates over time must therefore reflect a concomitant increase in the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater from -8$\permil$ in the Cambrian to its present value of $\sim$0$\permil$ \citep{Jaffres2007, Veizer1999}.

Carbonate clumped isotope thermometry supplies a unique insight into this problem because it provides data that can help resolve the two main complications of the carbonate $\delta^{18}$O proxy: the clumped isotope composition of a carbonate is independent of the $\delta^{18}$O of water and highly sensitive to diagenetic alteration at temperatures different from those of original deposition. The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates \citep{Eiler2011}, and isotopic clumping in a wide diversity of shallow-water calcifiers has been shown to record ambient water temperatures \citep{Ghosh2006, Henkes2013, Saenger2012, Tripati2010}. There is some uncertainty regarding the uniformity of these calibrations, but all previous studies suggest this phenomenon is a reliable and reasonably precise thermometer when calibrated against carbonates of the same type as analyzed in the same lab. Clumped isotope temperature estimates can in turn be used, with the $\delta^{18}$O of shell calcite and the experimentally determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}, to calculate the $\delta^{18}$O of the water from which the calcite was precipitated.

Fossils that have been deeply buried in lithified sediments can experience alteration at temperatures greater than those of original deposition. This alteration is often associated with recrystallization, which can change the $\delta^{18}$O and trace-metal content of carbonate mineral phases but is frequently difficult to detect. Clumped isotopes are sensitive to this type of alteration because the recrystallized carbonate will take on a new clumped isotope temperature that can be much higher than the temperature of the pristine fossil calcite. If recrystallized material is sampled along with pristine fossil material during analysis (e.g. cements filling void space within a fossil), the clumped isotope temperature of the sample will be significantly higher than the temperature of the seawater where the organism lived. We demonstrate here that carbonate clumped isotopes can provide a more sensitive indicator of diagenetic recrystallization than the $\delta^{18}$O or the trace metal content of a given carbonate phase, particularly in rocks that undergo nearly rock-buffered diagenesis.

In carbonate strata that have been deeply buried and subjected to high temperatures for extended periods of time, diagenetic alteration can also take the form of closed-system diffusive reordering of atoms within the mineral lattice \citep{Dennis2013, Huntington2011, Quade2013, VanDeVelde2013}. During this process, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the measured temperature of the calcite. The amount of diffusion that occurs is controlled by both the burial temperature and the length of time for which the sample remains at this temperature. The effects of closed-system diffusive reordering can be difficult to recognize, because the reordering does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and re-precipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}. One way to recognize diffusive reordering is through its effects on the distribution of temperatures in a sample suite; diffusive reordering raises the mean temperature without changing the variance in temperature. In addition, the degree of diffusive reordering can be constrained by independent indicators of any given sedimentary succession’s time-temperature history, such as the conodont color alteration index (CAI) or rock evaluation parameters of organic matter maturity. Recent studies of closed-system diffusive reordering suggest the time-temperature conditions required to increase the clumped isotope temperature of a sample by more than the measurement precision of $\sim$1$^{\circ}$C are more than sufficient to change the color of conodont elements, raising the CAI above the pristine value of one \citep{Dennis2010, Epstein1976, Passey2012, Stolper2014}. Therefore, in sample suites with CAI=1 or pre-oil window catagenesis, elevated clumped isotope temperatures can be reasonably interpreted to be the result of diagenetic alteration due to recrystallization.

In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotope data, when combined with bulk O and C isotope analyses of the same materials, can be diagnostic of specific diagenetic processes and regimes (recrystallization vs. closed-system diffusive reordering; high vs. low water/rock ratio conditions; meteoric vs. marine pore fluids), and can be used to identify the most pristine end-member of a sample set \citep{Eiler2011}. Of particular interest for this study, as carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. During rock-buffered diagenesis, this causes the $\delta^{18}$O of the water to increase, while the $\delta^{18}$O of the altered carbonate is not affected. In contrast, during a water-buffered diagenetic regime, increasing temperature instead causes the $\delta^{18}$O of the altered carbonate to decrease. And, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the $\delta^{18}$O of both the carbonate and the water are diagnostically low. We demonstrate here that this distinction has important implications for the estimation of the sea surface temperature and $\delta^{18}$O of seawater in the past.

@@

Samples were collected from several sections exposed on the island of Gotland, Sweden that span the latest Llandovery through Wenlock epochs ($\sim$434 – 430 Ma) of the Silurian Period (Fig. 1). Ages of the formations and correlations between outcrops are constrained by conodont and graptolite biostratigraphy \citep{Jeppsson2006}. The sections record shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was constrained to be $19\pm5.1^{\circ}$ south of the equator during the Silurian Period \citep{Torsvik1992}.

The oldest exposed strata on Gotland comprise the Lower Visby Formation, which consists of marls interbedded with fine-grained limestones deposited below storm wave base (Fig. 2). Fossils are abundant throughout this unit \citep{Calner2004a, Samtleben1996}. The contact between the Lower and Upper Visby formations is marked by a condensed interval with abundant fossils of the large solitary rugose coral genus \textit{Phaulactis} \citep{Jeppsson1997, Jeppsson2006, Munnecke2003, Samtleben1996}. This contact occurs just above the Llandovery-Wenlock boundary as defined by conodont biostratigraphy \citep{Aldridge1993, Jeppsson1983, Mabillard1985}. We defined the \textit{Phaulactis} bed as a stratigraphic datum to align the sections, except those in the Slite Formation, which are measured relative to an arbitrary bed in the Slite Formation. The contact between the Lower and Upper Visby formations is also characterized by a facies change that represents a maximum flooding interval \citep{Calner2004b}. Carbonate content increases in the Upper Visby Beds, with a decreasing frequency of marl beds and the appearance of bioclastic limestones and reef mounds. Ripple bedforms and cross-stratification indicate deposition above storm wave base. The Upper Visby Formation is particularly fossiliferous \citep{Calner2004a, Samtleben1996}.

@@

The Silurian strata exposed on Gotland present attractive materials for a clumped isotope study. This section has long been recognized for its generally excellent fossil preservation. The strata on Gotland are very nearly flat lying, with a dip of less than 1$^{\circ}$, display only minimal evidence of tectonic disturbance and have not been deeply buried \citep{Calner2004a, Jeppsson1983}. Conodont fossils from this section yield a conodont color alteration index (CAI) of one \citep{Epstein1976, Jeppsson1983}. The Silurian strata on Gotland also have an abundance of fossils, especially brachiopods, throughout the section. We sampled fossil phases for clumped isotope measurements in order to target calcite that was first precipitated directly from seawater (as opposed to carbonate matrix, which may contain authigenic cements). Many studies of paleoclimate have targeted the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998, Samtleben2001}. The fibrous calcite fabric characteristic of the secondary layer has been well characterized in modern and fossil brachiopods \citep{PerezHuerta2007, Samtleben2001, Schmahl2004}, providing a standard of comparison for the samples in our dataset. The trace metal composition of brachiopods has also been extensively studied as an indicator of diagenetic alteration \citep{Azmy1998, Brand2003, Brand2012, Mii1994, Grossman1996, Shields2003}. The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often than brachiopods for paleoenvironmental studies, but have been examined by other clumped isotope studies of paleotemperature \citep{Came2007, Dennis2013, Finnegan2011}. Measurements of several fossil taxa allow comparison of the relative preservation of different calcitic materials and can help to detect taxon-specific ``vital effects'' if they are present.

A final attractive feature of the Silurian section on Gotland is the previously observed positive excursion in $\delta^{13}$C that initiates at the boundary between the Lower and Upper Visby Beds (Fig. 2). This excursion, termed the Ireviken event, is now recognized globally \citep{Munnecke2003}. In Gotland, a 3$\permil$ excursion in $\delta^{13}$C is coupled to a weakly resolved 0.4$\permil$ positive excursion in $\delta^{18}$O \citep{Munnecke2003}. It is somewhat surprising that the positive excursions coincide with a maximum flooding interval, but it is difficult to distinguish the relative contributions of sediment supply and eustatic sea level rise to this facies change \citep{Calner2004a}. The excursions are also broadly coincident with minor extinctions in several benthic and pelagic marine taxa \citep{Munnecke2003}.

The coincident isotopic excursions and stratigraphic change suggest that the Ireviken Event captures a climate change phenomenon, the nature of which is still debated. The similarity between the isotopic signatures of the Ireviken event and known glaciations, such as the Late Ordovician glaciation, have led several studies to hypothesize a glacial cause for the Ireviken event \citep{Azmy1998, Kaljo2003, Brand2006, Calner2008}. This hypothesis was supported by the presence of glacial tillites in Brazil that were once interpreted to be early Wenlock in age \citep{Grahn1992}, but these diamictites were later re-interpreted as re-workings of earlier glacial deposits. Early Silurian glaciation in South America is currently thought to have ended prior to the late Llandovery, before the Ireviken Event \citep{DiazMartinez2007a, DiazMartinez2007b}. Many of the characteristics of the Ireviken Event, although not necessarily its global nature, could also be explained by a shift from a humid period to an arid period in the tropics, which could accompany a glaciation or occur independently \citep{Bickert1997, Jeppsson1990, Munnecke2003, Samtleben1996}. According to this model, during the humid period recorded by the Lower Visby Formation, increased precipitation would have lowered the $\delta^{18}$O of surface waters and also stimulated the upwelling of deep water with low $\delta^{13}$C onto the shelves. Increased precipitation would have caused intense weathering, which could explain the dominance of marls over carbonate reefs during this time. During the succeeding arid period beginning in the Upper Visby Formation, evaporation would have increased the $\delta^{18}$O of surface waters, and downwelling of saline water would have brought surface water with high $\delta^{13}$C onto the shelves. Low terrigenous influxes would have allowed the growth of carbonate reefs as observed in the H\"{o}gklint Formation. These hypotheses make different predictions for temperature and the oxygen isotopic composition of seawater. Here we test our ability to resolve possible changes in ocean temperature and/or seawater $\delta^{18}$O through the Ireviken event using stratigraphic samples from Gotland. Resolving a change this small with the uncertainties associated with the carbonate clumped isotope thermometer is challenging because a shift of 0.4$\permil$ in $\delta^{18}$O, if caused by change in temperature of carbonate precipitation at constant $\delta^{18}$O of water, corresponds to a change in temperature of approximately 2$^{\circ}$C. Though the uncertainties for a single carbonate clumped isotope temperature are commonly similar to or higher than this, replicate analyses can reduce these errors to $\pm$1$^{\circ}$C or even $\pm$0.4$^{\circ}$C \citep{Thiagarajan2011}.
